---
title: "test"
author: "Yunan Chen"
date: "2024-09-24"
output: pdf_document
---
# Reading 2

Since the evidence about specific carbohydrate diet (SCD) for inflammatory bowel disease (IBD) is remain limited, this study was conducted in the purpose of examining the impact of two different diets, SCD and modified SCD (MSCD) on IBD symptoms and inflammation
in childern and adolescents with IBD. 

To achieve this goal, an N-of-1 trail study design was performed. 54 individuals qualified for the inclusion criteria were first started with at least 1-week baseline on their usual diet (US). Then, randomized to begin either SCD or MSCD with a centralized, stratified, block randomization approach using a 1:1 allocation ratio. To avoid confounding the secular trends with the real treatment effect, multiple crossovers was conducted, which means that participants alternated between SCD and MSCD for 4 periods, each 8 weeks long. 

Analysis was done on both individual and aggregated level. Results of individual level was presented in tables showing participants had no clinically meaningful difference between SCD and MSCD. Change in IBD symptoms on SCD versus UD varied by individual. On aggregated level, there was <1% probability of a clinically meaningful difference in IBD symptoms between SCD and MSCD across the pooled, imputed sample. And there was a 62% probability that SCD was better than UD for reducing IBD symptoms, and 45% probability that MSCD was better than UD.

# Reading Missing data
1. MCAR: MCAR, missing completely at random, is when the missing is completely random and it does not depend on any variables in the data. This means that the probability of missing data on a variable is unrelated to any observed or unobserved data. An example for MCAR is in a heart disease study, the technician accidently loses some lab samples of some participants. Here, the missingness is not related to any participants' health related or other observed and unobserved variables, making it completely random.

2. MAR: MAR, missing at random, is when the missingness is related to observed variables, but not on the missing variable itself. In this situation, the missingness can be explained by other observed variables, but not the missing variable itself. An example of MAR is, in a study of depression, male participants are less likely to complete the survey questions asking about depression severity compared to the female participants, leading to missingness. This missingness is not related to the severity of participants' severity of depression, but is related to their gender which is observed.

3. MNAR: MNAR, missing not at random, is when the probability of missing data is related to the incompleted variables itself. An example for MNAR is that people with higher incomes may be more likely to leave the income question blank due to privacy concerns in a survey. Here, the missing of income response is related to participants' income level but not related to other observed variables.

In a longitudinal depression study, people complete surveys regularly through time and some participants might drop out from the study. This missingness might have various causes which might be difficult to distinguish among these three types. For example, if the participants move to other places where the reason for the moving is not related to any variables in our studies, this missingness is random and it should be MCAR. Or, if the participants are too old to make regular appointment and complete the survey, the missing data is related to their age, but not related to their severity of depression. This case is MAR. Or, if the participants' depression severity becomes much worse and they drop out since they are not willing to share their worse conditions, this missingness is MNAR where it is related to the missing variable itself. In this case, it would be difficult to tell which setting is appropriate for our data because in reality, researchers might not have clear information about the drop our reason and drop out might occur due to mix of reasons mentioned earlier.


Multiple imputation is a modern statistical technique to handle missing data. The basic procedure for multiple imputation includes three steps: the imputation phase, the analysis phase, and the pooling phase. In the imputation phase, we generate a specified number of datasets with different estimates of missing values. Then, we perform analyses separately using the same methods and steps (with no change due to missingness) on each dataset generated during the imputation phase, yielding multiple sets of estimates of each parameter and standard error. Last, in the pooling phase, we combine these multiple sets of results from each dataset's analysis into a single set of results.

Multiple imputation might not be appropriate to use when MNAR occurs, that is, when the missingness is related to the missing variable itself. This is because multiple imputation assumes the missingness in the data can be explained by observed data. The implementation of multiple imputation on MNAR data would lead to biases. Or, if our data has small sample size, it would be challenged to do reliable imputation using limited observations. Moreover, if the proportion of missingness is extremely high, our imputation might also be unreliable due to lack of information which would introduce uncertainty.



# EDA
```{r}
library(tidyr)
library(gtsummary)
schmid_data <- read.csv("schmid_data.csv")
schmid_trach <- schmid_data %>%
  select(mat_race, Trach) %>%
  tbl_summary(by=Trach,
              label = list(mat_race ~ "Race")) %>%
  modify_header(update = list(stat_1 ~ "**No**",
                              stat_2 ~ "**Yes**")) %>%
  modify_spanning_header(update =  all_stat_cols() ~  "**Trachoestomy**") %>%
  modify_footnote(update = all_stat_cols() ~ "median (IQR) for continuous; n (%) for categorical")
schmid_trach

schmid_death <- schmid_data %>%
  select(mat_race, Death) %>%
  tbl_summary(by=Death,
              label = list(mat_race ~ "Race")) %>%
  # modify_header(update = list(stat_1 ~ "**No**",
  #                             stat_2 ~ "**Yes**")) %>%
  modify_spanning_header(update =  all_stat_cols() ~  "**Death**") %>%
  modify_footnote(update = all_stat_cols() ~ "median (IQR) for continuous; n (%) for categorical")
schmid_death

schimd_tbl <- tbl_merge(list(schmid_trach, schmid_death),
                        tab_spanner = c("**Trachoestomy**", "**Death**"))
schimd_tbl
```

# Multiple Imputation
```{r}
library(HDSinRdata)
library(mice)
library(dplyr)
library(ggplot2)
data(pain)

# Missing Patterns
## Step 1
## The code below calculates whe percent of data is missing. A simple look at this table warns us about several variables that have more than 25% missing - such as PAIN_INTENSITY_AVERAGE.FOLLOW_UP, PROMIS_PAIN_BEHAVIOR, and BMI. It might be useful to remove these varibales from the dataset first as they might mess up the imputation.

p_missing <- unlist(lapply(pain, function(x) sum(is.na(x))))/nrow(pain)
sort(p_missing[p_missing > 0], decreasing = TRUE)

# Multiple Imputation
## Step 3
pain_mod <- pain[, -c(1:75)] %>%
  select(-PAIN_INTENSITY_AVERAGE.FOLLOW_UP)

pain_mod$PAT_SEX <- as.factor(pain_mod$PAT_SEX)
pain_mod$PAT_RACE <- as.factor(pain_mod$PAT_RACE)
pain_mod$CCI_BIN <- as.factor(pain_mod$CCI_BIN)
pain_mod$MEDICAID_BIN <- as.factor(pain_mod$MEDICAID_BIN)
## Step 4
pain_mice <- mice(pain_mod, maxit = 5, seed=(2550))

# Imputed Results
## Step 5
mice_imp1 <- mice::complete(pain_mice,1)
score1 <- mice_imp1 %>%
  group_by(PAIN_INTENSITY_AVERAGE) %>%
  summarize(
    avg_mental_health = mean(GH_MENTAL_SCORE),
    se_mental_health = sd(GH_MENTAL_SCORE)/sqrt(n()),
    avg_physical_health = mean(GH_PHYSICAL_SCORE),
    se_physical_health = sd(GH_PHYSICAL_SCORE)/sqrt(n())
  )

## Step 6
mice_imp2 <- mice::complete(pain_mice,2)
score2 <- mice_imp2 %>%
  group_by(PAIN_INTENSITY_AVERAGE) %>%
  summarize(
    avg_mental_health = mean(GH_MENTAL_SCORE),
    se_mental_health = sd(GH_MENTAL_SCORE)/sqrt(n()),
    avg_physical_health = mean(GH_PHYSICAL_SCORE),
    se_physical_health = sd(GH_PHYSICAL_SCORE)/sqrt(n())
  )


mice_imp3 <- mice::complete(pain_mice,3)
score3 <- mice_imp3 %>%
  group_by(PAIN_INTENSITY_AVERAGE) %>%
  summarize(
    avg_mental_health = mean(GH_MENTAL_SCORE),
    se_mental_health = sd(GH_MENTAL_SCORE)/sqrt(n()),
    avg_physical_health = mean(GH_PHYSICAL_SCORE),
    se_physical_health = sd(GH_PHYSICAL_SCORE)/sqrt(n())
  )


mice_imp4 <- mice::complete(pain_mice,4)
score4 <- mice_imp4 %>%
  group_by(PAIN_INTENSITY_AVERAGE) %>%
  summarize(
    avg_mental_health = mean(GH_MENTAL_SCORE),
    se_mental_health = sd(GH_MENTAL_SCORE)/sqrt(n()),
    avg_physical_health = mean(GH_PHYSICAL_SCORE),
    se_physical_health = sd(GH_PHYSICAL_SCORE)/sqrt(n())
  )

mice_imp5 <- mice::complete(pain_mice,5)
score5 <- mice_imp5 %>%
  group_by(PAIN_INTENSITY_AVERAGE) %>%
  summarize(
    avg_mental_health = mean(GH_MENTAL_SCORE),
    se_mental_health = sd(GH_MENTAL_SCORE)/sqrt(n()),
    avg_physical_health = mean(GH_PHYSICAL_SCORE),
    se_physical_health = sd(GH_PHYSICAL_SCORE)/sqrt(n())
  )

mice_pool<- rbind(score1, score2, score3, score4, score5)
mice_pool
score_pool <- mice_pool %>%
  group_by(PAIN_INTENSITY_AVERAGE) %>%
  summarize(
    avg_mental_health_pool = mean(avg_mental_health),
    
    se_mental_health_pool = sqrt(mean(se_mental_health^2)
    +(sum((avg_mental_health-mean(avg_mental_health))^2))/4
    +(sum((avg_mental_health-mean(avg_mental_health))^2))/20),
    
    avg_physical_health_pool = mean(avg_physical_health),
    se_physical_health_pool = sqrt(mean(se_mental_health^2)
    + (sum((avg_physical_health-mean(avg_physical_health))^2))/4
    +(sum((avg_physical_health-mean(avg_physical_health))^2))/20)
    )

ggplot(score_pool, aes(x = factor(PAIN_INTENSITY_AVERAGE))) +
  geom_point(aes(y = avg_mental_health_pool, color = "Mental Health")) +
  geom_line(aes(y = avg_mental_health_pool, color = "Mental Health"), group=1) +
  geom_errorbar(aes(ymin = avg_mental_health_pool - se_mental_health_pool,
                  ymax = avg_mental_health_pool + se_mental_health_pool, color = "Mental Health",), alpha = 0.6) +
  
  geom_point(aes(y = avg_physical_health_pool, color = "Physical Health")) +
  geom_line(aes(y = avg_physical_health_pool, color = "Physical Health"), group=1) +
  geom_errorbar(aes(ymin = avg_physical_health_pool - se_physical_health_pool,
                  ymax = avg_physical_health_pool + se_physical_health_pool, color = "Physical Health"), alpha = 0.6) +
  labs(x = "Pain Intensity", y = "Average Score", title = "Average Mental and Physical Health by Pain Intensity with SE", color = "Score Type") +
  #scale_color_manual(values = c("Mental Health" = "red", "Physical Health" = "blue"))
  theme_minimal()

# PAIN_INTENSITY_AVERAGE
# PROMIS_PHYSICAL_FUNCTION
# GH_MENTAL_SCORE

# sex / bin
# variable missing together / amount of missing a follow-up
# goal: how well can we predict the combined outcome at different time points (birth/36: both data /44)
# prediction better over time (but 44 loosing info discharged) flow chart 
# three imputation, three population (subsetting birth /discharge >36 / discharge >44) benefit (missing pattern observed): reducing the aount of the missing data, don't want to use 44 data when considering data at 36 wk
```